Increasing prevalence of plant-fungal symbiosis across two centuries of environmental change

### 2024-05-02 Manuscript/endo_herbarium.tex
Update endo_herbarium.tex

	\section*{Abstract}
Species' distributions and abundances are shifting in response to climate change. 
Most species harbor microbial symbionts that have the potential to influence these responses.
Mutualistic microbial symbionts may provide resilience to environmental change by protecting their hosts from increasing stress. Alternatively, environmental change disrupts these interactions may lead to declines in hosts or symbionts. 
Microbes can be preserved within the dried plant tissue of herbarium specimens presenting an opportunity to quantify changes across broad temporal and spatial scales. 
We ask how the prevalence of symbionts of grasses, \emph{Epichloë} fungal endophytes, which can protect hosts from drought, have changed over time in response to climate change, and how these changes vary across hosts' ranges.
Specifically, we took seed samples from herbarium specimens of three grass host species collected over the last two centuries, quantified fungal hyphae presence within the seeds using microscopy, and evaluated spatial and temporal trends. 
Overall, endophytes have increased in prevalence over the last two centuries from 25\% prevalence to 75\% prevalence on average across the three hosts.
We also found that changes in prevalence were associated with observed changes in annual and seasonal climate drivers corresponding with each hosts' peak growing season. 
Thus we provide novel evidence for a cryptic biological response to climate change that may contribute to the resilience of the symbiosis through ongoing environmental change.

@@

Understanding how biotic interactions are altered by global climate change is a major goal of basic ecological research and conservation management \cite{gilman2010framework,blois2013climate}.
Documented responses to environmental change, such as shifts in species' distributions \cite{aitken2008adaptation} and phenology \cite{piao2019plant}, are typically blind to concurrent changes in associated biotic interactions.
Empirically evaluating these biotic changes, whether interacting species shift in tandem with their partners or not \cite{hillerislambers2013will}, is crucial to predicting future climate change responses. 
Such evaluations have been limited because data on species interactions are limited, with few datasets extending over the long time scales of contemporary climate change \cite{poisot2021global}.

Natural history collections, which were originally collected to study and preserve taxonomic diversity, present a unique opportunity to explore long-term changes in ecological interactions across broad spatial scales \citep{meineke2018unrealized}. 
Natural history collections, built and maintained by the efforts of thousands of scientists, are invaluable time machines, primarily comprised of physical specimens of organisms  along with information about the time and place of their collection. 
These specimens are samples preserving physical legacies of ecological processes and species' interactions from dynamically changing environments across time and space.
Researchers have used plant collections to document climate change responses, including shifts in phenology \citep{willis2017old, park2019herbarium,  berg2019examination}, as well as in rates of pollination \citep{pauw2011reconstruction, duan2019century}, and herbivory \citep{meineke2019herbarium}. 
However, focus has been lacking on long-term changes in a particularly common type of biotic interaction, microbial symbioses.

Microbial symbionts are common to all macroscopic organisms and can have important effects on their hosts' survival, growth and reproduction \cite{rodriguez2009fungal,mcfall2013animals}.
Many microbial symbionts act as mutualists, engaging in reciprocally beneficial interactions with their hosts that ameliorate environmental stress. 
For example, bacterial symbionts of insects, such as \emph{Wolbachia}, can improve their hosts' thermal tolerance \citep{truitt2019wolbachia, renoz2019evolutionary}, and arbuscular mycorrhizal fungi, documented in 70-90\% of families of land plants \citep{parniske2008arbuscular}, allow their hosts to persist through drought conditions by improving water and nutrient uptake \citep{cheng2021elucidating}.
On the other hand, changes in the mean and variance of environmental conditions may disrupt microbial  mutualisms by changing the costs and benefits of the interaction for each partner, leading the interaction to deteriorate \citep{aslan2013mutualism}. 
Coral bleaching (the loss of symbiotic algae) due to temperature stress \citep{sully2019global} is perhaps the best known example, but this phenomenon is not unique to corals.
Lichens exposed to elevated temperatures experienced loss of photosynthetic function along with changes in the composition of their algal symbiont community \citep{meyer2022climate}.
How commonly and under what conditions microbial mutualisms deteriorate or strengthen under climate change remain unanswered questions, but previous work suggests that these alternative responses may depend on the intimacy and specialization of the interaction as well as the physiological tolerances of the mutualist partners \citep{toby2010mutualisms, warren2014mutualism, rafferty2015phenological}.

Understanding of how microbial symbioses are affected by climate change is additionally complicated by spatial hetereogeneity in the direction and magnitude of environmental change \cite{ipcc_2021}. 
Beneficial symbionts are likely able to shield their hosts from environmental stress in locations that experience a small degree of change, but symbionts in locations that experience changes of large magnitude may be pushed beyond their physiological limits \cite{webster2008temperature}.
Additionally, symbionts are often unevenly distributed across their hosts' distribution.
Facultative symbionts may be absent from portions of the host range \cite{afkhami2014mutualist}, and hosts may engage with a diversity of partners (different species, or different locally-adapted strains of one species) across their environments \cite{frade2008variation, rolshausen2018quantifying}.
Identifying broader spatial trends in symbiont prevalence is therefore an important step in developing predictions for where to expect changes in the symbiosis in future climates.

\josh{
\emph{Epichloë} fungal endophytes are specialized symbionts of cool-season grasses, which have been documented in $\sim 30$\% of cool-season grasses \citep{leuchtmann1992systematics}
They are transmitted vertically from maternal plants to offspring through the grasses' seeds.
Vertical transmission creates a feedback between the fitness of host and symbiont \citep{fine1975vectors, douglas1998host, rudgers2009fungus}. 
Over time, endophytes that act as mutualists for their hosts should rise in prevalence within a population. 
\emph{Epichloë} are known to improve their hosts' drought tolerance \cite{decunta2021systematic}, and protect their hosts against herbivores \cite{crawford2010fungal} and pathogens \cite{xia2018role} likely through the production of a diverse suite of alkaloids and other secondary metabolites.
The fitness feedback induced by vertical transmission leads to the prediction that endophyte prevalence should be high in populations where these fitness benefits are most important. 
Previous survey studies have documented large-scale spatial patterns in  endophyte prevalence structured by environmental gradients \citep{granath2007variation,bazely2007broad, afkhami2012fungal,sneck2017variation}.
We predicted that prevalence should track temporal changes in environmental drivers that elicit these fitness benefits.
For example, endophyte-mediated drought tolerance should lead prevalence to increase in regions where precipitation declines over time.
Early research on \emph{Epichloë} used herbarium specimens to describe the broad taxonomic diversity of hosts \citep{white1985endophyte}. 
Grasses are commonly identified and collected based on the presence of their reproductive structures meaning that collected specimens typically contain seeds, conveniently preserving the fungi along with their host plants on herbarium sheets. 
This creates the opportunity to leverage the unique spatio-temporal sampling of herbarium collections to examine the response of the symbiosis to historical climate change.
}{Volker's suggestions included moving all material specific to study system into methods, but it seems a bit difficult to do.}

In this study, we assessed for the first time the long-term responses of endophyte symbiosis to climate change through the use of herbarium specimens of three \emph{Epichloë} endophyte host species, \emph{Agrostis hyemalis}, \emph{Agrostis perennans}, and \emph{Elymus virginicus}.
We first address questions describing spatial and temporal trends in endophyte prevalence: (i) How has endophyte prevalence changed over time? and (ii) How spatially heterogenous are temporal trends in endophyte prevalence?
We then address how climate change may be driving these trends in endophyte prevalence by asking: (iii) What is the relationship between variation in temporal trends in endophyte prevalence and changes in climate drivers?
We examined 2,346 specimens collected across across eastern North America between 1824 and 2019.

\section*{Methods}
        \subsection*{Focal species}
Our surveys focused on three endophyte hosts, \emph{Agrostis hyemalis}, \emph{Agrostis perennans}, and \emph{Elymus virginicus}. Both \emph{Agrostis} species host \emph{Epichloë amarillans} \cite{craven2001multigene, leuchtmann2014nomenclatural}, while \emph{Elymus virginicus} typically hosts \emph{Epichloë elymi} \cite{clay2002evolutionary}.
These $C_3$ grass species are commonly represented in natural history collections with broad distributions covering much the Eastern United States.
\emph{A. hyemalis} is a small short-lived perennial species that germinates in the spring and typically blooms between March and July (most common collection month: May).
\emph{A. perennans} is of similar stature but is longer lived than \emph{Agrostis hyemalis} and blooms in late Summer and early Autumn (most common collection month: September). 
This species is relatively more sparsely distributed, tending to be found in shadier and more moist habitats while \emph{A. hyemalis} is often found in open and recently disturbed ground. 
Both \emph{Agrostis} species are recorded from throughout the Eastern US, but \emph{A. perennans} has a slighty more northern distribution, whereas \emph{A. hyemalis} is found rarely as far north as Canada and is listed as a rare plant in Minnesota.
\emph{E. virginicus} is a larger and relatively longer-lived  species that is more broadly distributed than the \emph{Agrostis} species. 
It begins flowering as early as March or April but continues throughout the Summer (most common collection month: July).

		\subsection*{Herbarium surveys}
We visited nine herbaria between 2019 and 2022 (see Table A1 for a summary of specimens included from each collection). 
In total, we quantified endophyte symbiosis for $1135$ \emph{A. hyemalis} specimens collected between 1824 and 2019, $357$ \emph{A. perennans} specimens collected between 1863 and 2017, and $854$ \emph{E. virginicus} specimens collected between 1839 and 2019 (Fig. 1, Fig 2A).

Our sampling plan was designed to minimize damage to these specimens.
We chose our focal species in part because they are commonly represented in herbarium collections, and produce high numbers of seeds, meaning that small samples would not diminish the value of the specimens for future studies. 
We collected 5-10 seeds per specimen after examining the herbarium sheet under a dissecting microscope to ensure that we collected mature seeds, not florets or unfilled seeds, fit for our purpose of identifying fungal endophytes with microscopy.
We excluded specimens for which information about the collection location and date were unavailable.
Each specimen was assigned geographic coordinates based on collection information recorded on the herbarium sheet using the geocoding functionality of the ggmap R package \citep{kahle2019package}.
Many specimens had digitized collection information readily available, but for those that did not, we transcribed information from pictures of the specimens. 
Collections were geo-referenced to the nearest county centroid, or nearest municipality when that information was available. 
For a few of the oldest specimens, only information at the state level was available, and so we used the state centoid.

After collecting seed samples, we quantified the presence or absence of \emph{Epichloë} fungal hyphae, which grow intercellularly, in each specimen using microscopy. 
We first softened seeds with a 10\% NaOH solution, then stained the seeds with aniline blue dye and squashed them under a microscope cover slip. 
We examined the squashed seeds for the presence of fungal hyphae at 100X magnification \cite{bacon2018stains}.
In some cases, the tissues examined during microscopy came from flowers or otherwise non-viable seeds, which were excluded from the seed counts for that specimen.
On average we scored $4.7$ seeds per specimen in \emph{A. hyemalis}, $4.2$ seeds per specimen in \emph{A. perennans}, and $3.8$ seeds per specimen in \emph{E. virginicus}.
Due to imperfect vertical transmission \cite{afkhami2008symbiosis}, it is possible that symbiotic host-plants produce a mixture of symbiotic and non-symbiotic seeds. 
We therefore designated each specimen as symbiotic if endophyte hyphae were observed in one or more seeds, or non-symbiotic if hyphae were observed in zero seeds. 
To capture uncertainty in the identification process, we recorded both a "liberal" and a "conservative" endophyte status for each plant.  
When we identified potential endophytes with unusual morphology, low uptake of stain, or a small amount of fungal hyphae across the scored seeds, we recorded a positive liberal status (more likely to be a true endophyte) and a negative conservative status (less likely to be a true endophyte). 
$89$\% of scored plants had matching liberal and conservative endophyte statuses, reflecting high confidence in endophyte identification.
The following analyses presented in the main text used the liberal statuses, but we repeated all analyses with the conservative statuses which yielded qualitatively similar results (Figure A5).

\begin{figure}[h]
	\centering
	\includegraphics[width = \linewidth]{collections_map.png}
	\caption{\textbf{Collection locations of herbarium specimens sampled for endophyte presence absence}. Color designates host species ( \emph{A. hyemalis} (green), \emph{A. perennans} (orange), \emph{E. virginicus} (purple))}
\end{figure}

\subsection*{Assessing spatial and temporal changes in endophyte prevalence}
We assessed spatial and temporal changes in endophyte prevalence across each host distribution, first quantifying the long-term overall trends and then examining how these trends vary across the spatial extent of each host distribution.
We used an approximate Bayesian method, Integrated Nested Laplace Approximation (INLA), to construct spatio-temporal models of endophyte prevalence.
INLA provides a computationally more efficient method of ascertaining model posteriors for certain models that can be formulated as latent Gausssian Models \cite{rue2009approximate}. 
Many common statistical models, including structured and unstructured mixed-effects models, can be represented as latent Gaussian Models. 
Fitting models with structured spatial effects is possible with MCMC sampling but can require long computation times, making INLA an effective alternative, which has been used to model spatial patterns in flowering phenology \cite{willems2022forest}, the abundance of bird species \cite{meehan2019spatial} and butterflies cite{crossley2022opposing}, the distribution of temperate trees \cite{engel2022spatial} as well as the population dynamics of endangered amphibians \cite{knapp2016large} and other ecological processes \cite{beguin2012hierarchical}.

To quantify overall trends in endophyte prevalence, we modeled endophyte presence/absences ($P$)  as a Bernoulli response variable for each specimen $s$ at a given location $i$.
In all cases, models for each species were fit separately.

\begin{subequations}
	\label{eq:trends}
	\begin{align}
		P_{s.i} \sim Bernoulli(\hat{P_{s,i}}) \\
		logit(\hat{P}_{s,i}) = \alpha_{s,i} + \beta_{s}*year_i+
		\chi_{i} + \omega_{i} 
	\end{align}
\end{subequations}

The expected endophyte prevalence, $\hat{P}$, was modelled with spatially-indexed random intercepts $\alpha$ that accounts for potential spatial autocorrelation between data points, and with a slope $\beta$ describing the overall change in endophyte prevalence. 
We accounted for potential biases introduced during the process of collecting specimens as well as in scoring ability by including random effects specific to each collector $\chi$ and to each scorer $\omega$.
Previous work has questioned whether the behavior of historical botanists and uneven sampling may introduce biases into ecological inferences made from historic collections \cite{kozlov2020biases}. 
Prolific collectors who contribute thousands of specimens may be more or less likely to collect certain species, or specimens with certain traits \cite{daru2018widespread}. 
Similarly, the process of scoring seeds for hyphae involved many individual researchers, who may vary in their likelihood of positively identifying fungal hyphae. 
By including a random effect for collectors and for scorers, we accounted for variance across individual researchers that may bias our predictions of changes in endophyte prevalence.

To quantify how temporal trends may vary spatially, we repeated the modelling above, but instead incorporating a spatially-varying coefficient for collection year.

\begin{subequations}
	\label{eq:trends}
	\begin{align}
		P_{s.i} \sim Bernoulli(\hat{P_{s,i}}) \\
		logit(\hat{P}_{s,i}) = \alpha_{s,i} + \tau_{s,i}*year_i +
		\chi_{i} + \omega_{i} 
	\end{align}
\end{subequations}

Note that the expected endophyte prevalence, $\hat{P}$, was modelled as above, and that $\tau$ describes the spatially-varying random slope for collection year.
This spatially-varying random effect allowed us to flexibly estimate variation in slope at locations across the study region.
All spatially-structured random intercepts and slopes were constructed using stochastic partial differential equations (SPDE) that depend on a covariance matrix according to the proximity of each collection location \citep{lindgren2011explicit,bakka2018spatial}. 
The covariance matrix was approximated using a Matérn covariance function, with each data point assigned a location according to the nodes of a mesh of non-overlapping triangles across our study area (Fig A2).

We performed model fitting using the inlabru R package \citep{}, with vague priors, and compared models with different sizes of mesh, which had little effect on the resulting model estimates.
Each mesh was bounded by the predicted host distribution, described below.
Posterior modes were stable indicating that numeric convergence was successful.
We assessed model fit with graphical posterior predictive checks (Fig. A3).
The model performed adequately at classifying the historical data, comparing the accuracy of predictions from the model with observed data (avg. AUC = 0.77; Fig. A4).

@@

\begin{figure}[H]
	\centering
	\includegraphics[width = \linewidth]{year_plot.png}
	\caption[Temporal trends in endophyte prevalence.]{Temporal trends in endophyte prevalence. (A) Histograms show the frequency of collection through time for each host species. (B) Colored points are binned means of the observed endophyte presence/absence data (black dashes). Colors represent each host species and point size is determined by the number of specimens. Lines show predicted mean endophyte prevalence over the study period along with the 50\%  and 95\% CI bands.}
\end{figure}

>>>>

	\section*{Abstract}
Species' distributions and abundances are shifting in response to climate change. 
Most species harbor microbial symbionts that have the potential to influence these responses.
Mutualistic microbial symbionts may provide resilience to environmental change by protecting their hosts from increasing stress. 
Alternatively, environmental change that disrupts these interactions may lead to declines in hosts or symbionts. 
Microbes preserved within herbarium specimens offer a unique opportunity to quantify changes in microbial symbiosis across broad temporal and spatial scales. 
We asked how the prevalence of seed-transmitted fungal symbionts of grasses (\emph{Epichloë} endophytes), which can protect hosts from abiotic stress, have changed over time in response to climate change, and how these changes vary across host species' ranges.
Specifically, we analyzed \tom{\#}{Give numbers} herbarium specimens of three grass host species collected over the last two centuries (18\#\# -- 20\#\#) for the presence or absence of endophyte symbiosis, and evaluated spatial and temporal trends in endophyte prevalence. 
We found that endophytes have increased in prevalence over the last two centuries from ca. 25\% prevalence to ca. 75\% prevalence, on average, across the three host species.
We also found that changes in prevalence were associated with observed \tom{changes in annual and seasonal climate drivers}{Describe ``changes'' -- warming? drying?} corresponding to each host species' peak growing season. 
Our results provide novel evidence for a cryptic biological response to climate change that may contribute to the resilience of host-microbe symbiosis through context-dependent benefits that confer a fitness advantage to symbiotic hosts under environmental change.

@@

Understanding how biotic interactions are altered by global change is a major goal of basic and applied ecological research \cite{gilman2010framework,blois2013climate}.
Documented responses to environmental change, such as shifts in species' distributions \cite{aitken2008adaptation} and phenology \cite{piao2019plant}, are typically blind to concurrent changes in associated biotic interactions.
Empirically evaluating these biotic changes -- whether interacting species shift in tandem with their partners or not \cite{hillerislambers2013will} -- is crucial to predicting the reogranization of Earth's biodiversity under global change. 
Such evaluations have been limited because few datasets on species interactions extend over sufficiently long time scales of contemporary climate change \cite{poisot2021global}.

Natural history specimens, which were originally collected to study and preserve taxonomic diversity, present a unique opportunity to explore long-term changes in ecological interactions across broad spatial and temporal scales \citep{meineke2018unrealized}. 
Natural history collections, built and maintained by the efforts of thousands of scientists, are invaluable time machines, primarily comprised of physical specimens of organisms along with information about the time and place of their collection. 
These specimens often preserve physical legacies of ecological processes and species' interactions from dynamically changing environments across time and space.
For example, previous researchers have used plant collections (herbaria) to document shifts in phenology \citep{willis2017old, park2019herbarium,  berg2019examination}, pollination \citep{pauw2011reconstruction, duan2019century}, and herbivory \citep{meineke2019herbarium} related to anthropogenic climate change. 
However, few previous studies have leveraged biological collections to examine climate change-related shifts in a particularly common type of interaction: microbial symbiosis.

Microbial symbionts are common to all macroscopic organisms and can have important effects on their hosts' survival, growth and reproduction \cite{rodriguez2009fungal,mcfall2013animals}.
Many microbial symbionts act as mutualists, engaging in reciprocally beneficial interactions with their hosts that can ameliorate environmental stress. 
For example, bacterial symbionts of insects, such as \emph{Wolbachia}, can improve their hosts' thermal tolerance \citep{truitt2019wolbachia, renoz2019evolutionary}, and arbuscular mycorrhizal fungi, documented in 70-90\% of families of land plants \citep{parniske2008arbuscular}, allow their hosts to persist through drought conditions by improving water and nutrient uptake \citep{cheng2021elucidating}.
On the other hand, changes in the mean and variance of environmental conditions may disrupt microbial  mutualisms by changing the costs and benefits of the interaction for each partner, leading the interaction to deteriorate \citep{aslan2013mutualism}. 
Coral bleaching (the loss of symbiotic algae) due to temperature stress \citep{sully2019global} is perhaps the best known example, but this phenomenon is not unique to corals.
Lichens exposed to elevated temperatures experienced loss of photosynthetic function along with changes in the composition of their algal symbiont community \citep{meyer2022climate}.
How commonly and under what conditions microbial mutualisms deteriorate or strengthen under climate change remain unanswered questions.
Previous work suggests that these alternative responses may depend on the intimacy and specialization of the interaction as well as the physiological tolerances of the mutualist partners \citep{toby2010mutualisms, warren2014mutualism, rafferty2015phenological}.

Understanding of how microbial symbioses are affected by climate change is additionally complicated by spatial hetereogeneity in the direction and magnitude of environmental change \cite{ipcc_2021}. 
Beneficial symbionts are likely able to shield their hosts from environmental stress in locations that experience a small degree of change, but symbionts in locations that experience changes of large magnitude may be pushed beyond their physiological limits \cite{webster2008temperature}.
Additionally, symbionts are often unevenly distributed across their hosts' distribution.
Facultative symbionts may be absent from portions of the host range \cite{afkhami2014mutualist}, and hosts may engage with a diversity of partners (different symbiont species or locally-adapted strains) across their environments \cite{frade2008variation, rolshausen2018quantifying}.
Identifying broader spatial trends in symbiont prevalence is therefore an important step in developing predictions for where to expect changes in the symbiosis in future climates.

\emph{Epichloë} fungal endophytes are specialized symbionts of cool-season grasses, which have been documented in $\sim 30$\% of cool-season grass species \citep{leuchtmann1992systematics}.
They are transmitted vertically from maternal plants to offspring through seeds.
Vertical transmission creates a feedback between the fitness of host and symbiont \citep{fine1975vectors, douglas1998host, rudgers2009fungus}. 
Over time, endophytes that act as mutualists should rise in prevalence within a host population \citep{donald2021context}. 
\emph{Epichloë} are known to improve their hosts' drought tolerance \cite{decunta2021systematic} and protect their hosts against herbivores \cite{crawford2010fungal} and pathogens \cite{xia2018role} likely through the production of a diverse suite of alkaloids and other secondary metabolites.
The fitness feedback induced by vertical transmission leads to the prediction that endophyte prevalence should be high in populations where these fitness benefits are most important. 
Previous survey studies have documented large-scale spatial patterns in  endophyte prevalence structured by environmental gradients \citep{granath2007variation,bazely2007broad, afkhami2012fungal,sneck2017variation}.
We predicted that prevalence should track temporal changes in environmental drivers that elicit these fitness benefits.
%For example, endophyte-mediated drought tolerance should lead endophyte prevalence to increase in regions where precipitation has declined over time.

Early research on \emph{Epichloë} used herbarium specimens to describe the broad taxonomic diversity of host species that harbor these symbionts \citep{white1985endophyte}, \tom{establishing that signatures of endophyte symbiosis can be recovered from long-preserved plant tissue.}{Not sure about this but I am trying to pre-empt skepticism that fungi are less detectable in older samples.} 
However, no subsequent studies, to our knowledge, have used the vast resources of biological collections to quantitatively assess spatio-temporal trends in endophyte prevalence and their environmental correlates. 
Grasses are commonly collected and identified based on the presence of their reproductive structures, meaning that preserved specimens typically contain seeds, conveniently preserving the fungi along with their host plants on herbarium sheets. 
This creates the opportunity to leverage the unique spatio-temporal sampling of herbarium collections to examine the response of the symbiosis to historical climate change.

In this study, we assessed the long-term responses of endophyte symbiosis to climate change through the use of herbarium specimens of three North American host grass species (\emph{Agrostis hyemalis}, \emph{Agrostis perennans}, and \emph{Elymus virginicus}).
We first address questions describing spatial and temporal trends in endophyte prevalence: (i) How has endophyte prevalence changed over the past two centuries? and (ii) How spatially heterogenous are temporal trends in endophyte prevalence across eastern North America?
We then address how climate change may be driving trends in endophyte prevalence by asking: (iii) What is the relationship between variation in temporal trends in endophyte prevalence and changes in climate drivers?
We predicted that aggregate endophyte prevalence would increase over time in tandem with climate warming, and that hotspots of endophyte change would correspond spatially to hotspots of climate change. 
To answer these questions we examined a total of 2,346 specimens collected across eastern North America between 1824 and 2019.
\tom{}{I think the consensus was to keep the out-of-sample validation which should absolutely go into the Intro as an important element of novelty. Should go in the Asbtract too.}

\section*{Methods}
        \subsection*{Focal species}
Our surveys focused on three native North American grasses: \emph{Agrostis hyemalis}, \emph{Agrostis perennans}, and \emph{Elymus virginicus}. 
Both \emph{Agrostis} species host \emph{Epichloë amarillans} \cite{craven2001multigene, leuchtmann2014nomenclatural}, while \emph{Elymus virginicus} typically hosts \emph{Epichloë elymi} \cite{clay2002evolutionary}.
These $C_3$ grass species are commonly represented in natural history collections with broad distributions covering much the eastern United States.
\emph{A. hyemalis} is a small short-lived perennial species that germinates in the spring and typically flowers between March and July (most common collection month: May).
\emph{A. perennans} is of similar stature but is longer lived than \emph{Agrostis hyemalis} and flowers in late \tom{Summer}{Are seasons capitalized?} and early Autumn (most common collection month: September). 
\emph{A. perennans} is more sparsely distributed, tending to be found in shadier and more moist habitats, while \emph{A. hyemalis} is commonly found in open and recently disturbed ground. 
Both \emph{Agrostis} species are recorded from throughout the Eastern US, but \emph{A. perennans} has a slighty more northern distribution, whereas \emph{A. hyemalis} is found rarely as far north as Canada and is listed as a rare plant in Minnesota.
\emph{E. virginicus} is a larger and relatively longer-lived  species that is more broadly distributed than the \emph{Agrostis} species. 
It begins flowering as early as March or April but continues throughout the Summer (most common collection month: July).

		\subsection*{Herbarium surveys}
We visited nine herbaria between 2019 and 2022 (see Table A1 for a summary of specimens included from each collection). 
We permission from herbarium staff, we acquired seed samples from $1135$ \emph{A. hyemalis} specimens collected between 1824 and 2019, $357$ \emph{A. perennans} specimens collected between 1863 and 2017, and $854$ \emph{E. virginicus} specimens collected between 1839 and 2019 (Fig. \ref{fig:map}, Fig \ref{fig:temporal}A).
%Our sampling plan was designed to minimize damage to these specimens.
We chose our focal species in part because they are commonly represented in herbarium collections, and produce high numbers of seeds, meaning that small samples would not diminish the value of the specimens for future studies. 
We collected 5-10 seeds per specimen after examining the herbarium sheet under a dissecting microscope to ensure that we collected mature seeds, not florets or unfilled seeds, fit for our purpose of identifying fungal endophytes with microscopy.
We excluded specimens for which information about the collection location and date were unavailable.
Each specimen was assigned geographic coordinates based on collection information recorded on the herbarium sheet using the geocoding functionality of the ggmap R package \citep{kahle2019package}.
Many specimens had digitized collection information readily available, but for those that did not, we transcribed information printed on the herbarium sheet. 
Collections were geo-referenced to the nearest county centroid, or nearest municipality when that information was available. 
For a few of the oldest specimens, only information at the state level was available, and so we used the state centoid.

After collecting seed samples, we quantified the presence or absence of \emph{Epichloë} fungal hyphae, which grow intercellularly, in each specimen using microscopy. 
We first softened seeds with a 10\% NaOH solution, then stained the seeds with \tom{aniline blue}{Most of our stains in the lab also have lactic acid.} dye and squashed them under a microscope cover slip. 
We examined the squashed seeds for the presence of fungal hyphae at 200-400X magnification \cite{bacon2018stains}.
In some cases, the tissues examined during microscopy came from flowers or otherwise non-viable seeds, which were excluded for that specimen.
On average we scored $4.7$ seeds per specimen of \emph{A. hyemalis}, $4.2$ seeds per specimen of \emph{A. perennans}, and $3.8$ seeds per specimen of \emph{E. virginicus}; we scored \# seeds in total. .
Due to imperfect vertical transmission \cite{afkhami2008symbiosis}, it is possible that symbiotic host-plants produce a mixture of symbiotic and non-symbiotic seeds. 
We therefore designated a specimen as endophyte-symbiotic if \emph{Epichloë} hyphae were observed in one or more of its seeds, or non-symbiotic if hyphae were observed in none of its seeds. 
To capture uncertainty in the endophyte scoring process, we recorded both a "liberal" and a "conservative" endophyte status for each \tom{plant}{Plant or seed?}.  
When we identified potential endophytes with unusual morphology, low uptake of stain, or a small amount of fungal hyphae across the scored seeds, we recorded a positive liberal status (more likely to be endophyte-positive) and a negative conservative status (less likely to be endophyte-positive). 
$89$\% of scored plants had matching liberal and conservative scores, reflecting high confidence in endophyte status.
The following analyses in the main text used the liberal status, but we repeated all analyses with the conservative status which yielded qualitatively similar results \tom{(Figure A5)}{Use \ref{} for citing figures, and stay consistent between Figure and Fig.}.

\begin{figure}[h]
	\centering
	\includegraphics[width = \linewidth]{collections_map.png}
	\caption{\textbf{Collection locations of herbarium specimens of three grass host species across eastern North America that were sampled for \emph{Epichloë} endophyte presence or absence}.}
	\label{fig:map}
\end{figure}

\subsection*{Modeling spatial and temporal changes in endophyte prevalence}
We assessed spatial and temporal changes in endophyte prevalence across each host distribution, first quantifying the ``global'' trends, aggregating across space, and then examining spatial heterogeneity in the direction and magnitude of endophyte change (hotspots and coldspots) across the spatial extent of each host's distribution.
\tom{To appropriately account for the spatial non-independence of geo-referenced occurrences we used an approximate Bayesian method, Integrated Nested Laplace Approximation (INLA), to construct spatio-temporal models of endophyte prevalence.
INLA provides a computationally more efficient method of ascertaining parameter posterior distributions for certain models that can be formulated as latent Gausssian Models \cite{rue2009approximate}. 
Many common statistical models, including structured and unstructured mixed-effects models, can be represented as latent Gaussian Models. 
Fitting models with structured spatial effects is possible with MCMC sampling but can require long computation times, making INLA an effective alternative, which has been used to model spatial patterns in flowering phenology \cite{willems2022forest}, the abundance of bird species \cite{meehan2019spatial} and butterflies cite{crossley2022opposing}, the distribution of temperate trees \cite{engel2022spatial} as well as the population dynamics of endangered amphibians \cite{knapp2016large} and other ecological processes \cite{beguin2012hierarchical}.}{I think we need another sentence or two in this paragraph that provides an intuitive explanation for why INLA is the right approach for our analysis, written specifically for people who don't know what INLA is and don't want to know. I know the model includes a spatial decay term that controls spatial autocorrelation, a feature that I think is worth highlighting.}

First, to quantify global trends in endophyte prevalence, we modeled endophyte presence/absences of the $i^{th}$ specimen ($P_i$) as a Bernoulli response variable with expected probability of endophyte occurrence $\hat{P_i}$. 
We modeled $\hat{P_i}$ as a linear function of collection year and accounting for random effects associated with location ($l[i]$, a unique latitude-longitude combination), collector identity ($c[i]$), and scorer identity ($s[i]$) of the $i^{th}$ specimen.
\begin{subequations}
	\label{eq:trends}
	\begin{align}
		logit(\hat{P}_{i}) = \alpha_{l[i]} + \beta*year_i + \chi_{c[i]} + \omega_{s[i]} 
	\end{align}
\end{subequations}
Spatially-indexed random intercepts $\alpha_{l[i]}$ account for potential spatial autocorrelation between data points, and year slope $\beta$ describes the overall temporal trend in endophyte prevalence. 
We accounted for potential biases introduced during the process of collecting specimens as well as in scoring ability by including random effects specific to each collector $\chi$ and scorer $\omega$.
Previous work suggests that behavior of historical botanists and uneven sampling may introduce biases into ecological inferences made from historic collections \cite{kozlov2020biases}. 
Prolific collectors who contribute thousands of specimens may be more or less likely to collect certain species, or specimens with certain traits \cite{daru2018widespread}. 
Similarly, the process of scoring seeds for hyphae involved many student researchers who, even with standardized training, may vary in their likelihood of positively identifying \emph{Epichloë} hyphae. 
By including a random effect for collectors and for scorers, we accounted for variance across individual researchers that may bias our predictions of changes in endophyte prevalence.
\tom{Models for each host species were fit separately.}{It would be great to pull all species into one model and have them share variance terms for the random effects. I suspect such a model would give better and more stable estimates.}
\tom{}{I updated the notation in ways that make more sense to me, but you should check that this is true to the actual model (I think it is). Also, a more complete presentation of this model would show the variance terms for $\alpha$, $\chi$, and $\omega$. I presume the latter two are Gaussian but I don't know how to represent the distribution of $\alpha$.}

Second, to quantify how temporal trends may vary spatially, we repeated the modelling above, but incorporated a spatially-varying coefficient for collection year:
\begin{subequations}
	\label{eq:trends}
	\begin{align}
		logit(\hat{P}_{i}) = \alpha_{l[i]} + \beta_{l[i]}*year_i + \chi_{c[i]} + \omega_{s[i]} 
	\end{align}
\end{subequations}
The spatially-varying year slope $\beta_{l}$ allowed us to flexibly estimate variation in the temporal trajectory of endophyte change at locations across the study region.

\tom{For both models, spatially-structured random intercepts ($\alpha_{l}$) and slopes ($\beta_{l}$) were constructed using stochastic partial differential equations (SPDE) that depend on a covariance matrix according to the proximity of each collection location \citep{lindgren2011explicit,bakka2018spatial}. 
The covariance matrix was approximated using a Mat\'{e}rn covariance function, with each data point assigned a location according to the nodes of a mesh of non-overlapping triangles across our study area (Fig A2).}{This paragraph would be a place to describe the variance terms for the other random effects.}

We performed model fitting using the inlabru R package \citep{}, \tom{with vague priors}{I thought you needed informative priors on the spatial decay parameters}, and compared models with different sizes of mesh, \tom{which had little effect on the resulting model estimates}{That'sa good but you still need to state what mesh size you used and what that means, biologically.}.
Each \tom{mesh}{You have not defined what you mean by ``mesh''.} was bounded by the predicted host distribution, described below.
Posterior modes were \tom{stable}{Assessed how?} indicating that numeric convergence was successful.
We assessed model fit with graphical posterior predictive checks (Fig. A3).
\tom{The model performed adequately at classifying the historical data, comparing the accuracy of predictions from the model with observed data (avg. AUC = 0.77; Fig. A4). }{Maybe move this to validation section, and then have both in-sample and out-of-sample approaches.}

@@

\begin{figure}[H]
	\centering
	\includegraphics[width = \linewidth]{year_plot.png}
	\caption[Temporal trends in endophyte prevalence.]{Temporal trends in endophyte prevalence. (A) Histograms show the frequency of collection through time for each host species. (B) Colored points are binned means of the observed endophyte presence/absence data (black dashes). Colors represent each host species and point size is determined by the number of specimens. Lines show predicted mean endophyte prevalence over the study period along with the 50\%  and 95\% CI bands.}
	\label{fig:temporal}
\end{figure}
